---
title: "DH_Bio720_Assignment3"
output: html_document
---




```{r}

#read in csv 
rna_counts <- read.csv("https://www.dropbox.com/s/gqu520cc6r7xjw4/eXpress_dm_counts.csv?dl=1",  stringsAsFactors=FALSE)
#filter out genes with expression of 0
rna_counts <-  rna_counts[apply(rna_counts!=0, 1, all),]
#create row names 
rna_countsgood <- rna_counts[ , -1]
rownames(rna_countsgood) <- rna_counts[,1]
#create matrix - will be handy for some questions
rna_countsgood <- data.matrix(rna_countsgood, rownames.force = NA)

```


2) Create function that can calculate and output mean expression 
```{r}
#x argument is file and y argument is sample
mean2 <- function(x, y, log2scale=TRUE) {
  if (log2scale == TRUE) {
    log2(mean(x[ , y ]))
  } else {
    mean(x[, y])
  }
}

#show that function works with log2 and without
mean2(rna_counts, "F101_lg_female_hdhorn", log2scale = FALSE)
mean2(rna_counts, "F101_lg_female_hdhorn", log2scale = TRUE)

#show that function works with different sample 

mean2(rna_counts, "M120_sm_male_genitalia", log2scale = TRUE)

```


3) Use a loop to generate a vector of the mean expression value for each column 
```{r}
#change function to not require sample argument 

mean3 <- function(x, log2scale=TRUE) {
  if (log2scale == TRUE) {
    log2(mean(x))
  } else {
    mean(x)
  }
}

#create loop to generate mean expression for each column 
system.time(for (i in 2:56){
z <- mean3(rna_counts[ , i]) 
d <- colnames(rna_counts)[i]
print (paste(d,":", z))
}
)

#observe patterns 
sort(log2(colMeans(rna_countsgood)))

```
male_genitalia appears to have lower mean expression, compared to other tissues in the same males. In addition,  M180_lg_male_wings has the highest mean gene expression, while F196_sm_female_thxhorn has the lowest. I do not notice any major patterns for which individuals or tissues have the highest mean expression.

4)Repeat above but using apply family function 
```{r}

system.time(samplecounts <- sapply(rna_counts[, 2:56], mean3))
print(samplecounts)

```
A loop is slower than using the apply family function. 

5)What is an easier way to do this? 
```{r}

#use packages R has
colMeans(rna_countsgood)
log2(colMeans(rna_countsgood))

```

6) Write function to get mean expression of given gene across samples 

```{r}

#create function
mean4 <- function(x,y, log2scale=TRUE) {
  if (log2scale == TRUE) {
    log2(mean(x[ y, ]))
  } else {
    mean(x[ y,  ])
  }
}

#Show that function works
mean4(rna_countsgood, "FBpp0087248")

#Use apply to show each gene 
genecounts <- apply(rna_countsgood, 1, mean3)
print(genecounts)

```

7) Calculate the mean expression for the subset of columns for large and small male head horns

```{r}
#create subset of large male head horns
lg_male_hdhorn <- subset(rna_counts, select=c( "M125_lg_male_hdhorn", "M160_lg_male_hdhorn", "M180_lg_male_hdhorn","M257_lg_male_hdhorn"))

#create subset of small head horns 
sm_male_hdhorn <- subset(rna_counts, select=c("M120_sm_male_hdhorn",  "M171_sm_male_hdhorn", "M172_sm_male_hdhorn", "M200_sm_male_hdhorn"))

#create row names 
rownames(sm_male_hdhorn) <- rna_counts[,1]
rownames(lg_male_hdhorn) <- rna_counts[,1]

#gene counts
genecounts_sm_male_hdhorn <- apply(sm_male_hdhorn, 1, mean3, log2scale = FALSE)
genecounts_lg_male_hdhorn <- apply(lg_male_hdhorn, 1, mean3, log2scale = FALSE)

#gene counts log2 
genecounts_sm_male_hdhorn_log2 <- apply(sm_male_hdhorn, 1, mean3)
genecounts_lg_male_hdhorn_log2 <- apply(lg_male_hdhorn, 1, mean3)

#calculate difference 
difference <-(genecounts_lg_male_hdhorn - genecounts_sm_male_hdhorn)
differencelog2 <- (genecounts_lg_male_hdhorn_log2 - genecounts_sm_male_hdhorn_log2)

```


8) plot the mean expression of each gene and the difference in expression values
```{r}
#calculate avg gene expression
avgexp <- (genecounts_lg_male_hdhorn + genecounts_sm_male_hdhorn) / 2 
avgexplog2 <- (genecounts_lg_male_hdhorn_log2 + genecounts_sm_male_hdhorn_log2) / 2 


plot(avgexp, difference, main = "Average Gene Expression vs Difference", xlab = "Gene Expression", ylab = "Genes Expression Difference")


plot(avgexplog2, differencelog2, main = "Log2 Average Gene Expression vs Difference", xlab = "Avg Gene Expression", ylab = "Gene Expression Difference")

```


 

